Efficiently summarising relationships in large samples: a general duality between statistics of genealogies and genomes

### 2019-09-11 treestats_paper.tex
plr edits up until 1000G data

Many population genetic summary statistics can be defined using the allele frequency spectrum
of a collection of homologous sequences.
Concretely, write $p_i(a)$ for the frequency of allele $a$ at a site $i$,
and $h(p)$ for the allele frequency spectrum
-- i.e., the number of alleles whose frequency is $p$ in our sample --
and suppose that for some function $f()$, we want to compute $\sum_p h(p) f(p)$.
(For instance, mean genetic diversity is computed with $f(p) = p (1-p)$.)
Since this is a simple sum over sites,
we can compute this by summing over the $L$ loci, as
\begin{align*}
    \sum_p h(p) f(p) = \sum_{i=1}^L \sum_a f(p_i(a)).
\end{align*}
(Note that since the ancestral allele is included, this function is invariant to relabeling of alleles.)
If we know where each mutation has occurred on the genealogy,
we can easily find the frequency of the corresponding allele by counting how many samples
inherit from that mutation in the tree (and accounting for those that subsequently mutate).
Here, we only consider statistics whose values are invariant to monomorphic sites. Thus, for a sample of $n$
genomes, $0 < p \leq 1-1/n$.

What we do here is a generalization of this idea.
Roughly speaking, to do this we will
propagate ``weights'' additively up each tree,
summarize these weights on each branch with a ``summary function'',
and aggregate these summaries to produce statistics
(e.g., averaged in windows).
For instance, we could calculate how many of a certain set $A$ of samples
inherit from a particular edge in a tree
by (a) assigning each of these samples weight 1 (and other nodes weight zero), then
(b) finding the total weight of all nodes in the subtree inheriting from that edge.
It turns out to be useful to allow more general weights (including negative ones).
Figure \ref{fig:divergence_diagram} shows a simple example of these procedures
of propagating weights up a tree and summarising them,
used to calculate mean sequence divergence over a short sequence spanned by a single tree.

@@

\begin{definition}[Sample and subtree weights]
    The \emph{samples} of a tree sequence are the focal nodes,
    and appear in the trees at every point along the genome.
    A list of \emph{sample weights} $\iw$ assigns a numeric value $\iw(v)$
    to every sample node.
    Given these and a tree $T$,
    the \emph{subtree weight} $\nw_T(u)$ for a node $u$ is the sum of all sample weights
    of every sample node that is descended $u$ in the tree (including $u$, if it is a sample):
    \begin{align*}
        \nw_T(u) = \sum_{v \,:\, v \le_T u} \iw(v) ,
    \end{align*}
    where $v \le_T u$ if $u$ is on the path from $v$ to root in $T$.
\end{definition}

@@

We allow vector-valued weights,
i.e., $\iw(v)$ may be a vector $(\iw_1(v), \ldots, \iw_m(v))$,
so that when summarising we have access to more than one aspect of each subtree.
The requirement that $f(0) = f(\tiw) = 0$ is for consistency,
so that statistics do not depend on portions of the tree either not ancestral to any of the samples or ancestral to all
of them.
% Below, in section XXX, we describe how to efficiently
% maintain a list of the weights of the subtrees inherit from each node
% as we iterate through the tree sequence.

\begin{figure}
    \centering
    \includegraphics{figures/divergence_diagram}
    \caption{
        An example of computing sequence divergence between red nodes and blue nodes
        from seven mutations at distinct sites in a sequence of length 10 using the tree.
        The summary function, $f(a,b) = (a/3)(1-b/2)$ gives the probability that
        an allele found in $a$ of the red nodes and $b$ of the blue nodes is present in
        a randomly chosen red node but not in a randomly chosen blue node.
        \textbf{(a)} Weights are assigned to each subtree by adding together the weights of any children,
        plus the weight of the node itself, if it is a sample.
        \textbf{(b)} In this example, subtree weights $(a,b)$ record how many red nodes ($a$)
        and how many blue nodes ($b$) lie in that subtree.
        \textbf{(c)} The summary function shown is applied to both the weight of the subtree
        and the remaining weight of each mutation,
        so for instance the mutation above the node with weights $(2,1)$
        contributes $f(2,1) + f(1,1) = 1/3 + 1/6 = 1/2$.
        The final value for divergence is the sum of these values divided by 10,
        the total number of sites.
        \krt{This is a nice fig, but the $2 \times$ and the $/10 sites$ parts may be easy to miss?}
        \label{fig:divergence_diagram}
    }
\end{figure}

@@

Before defining the different classes of statistic,
it will help to establish a little more notation.
A tree sequence efficiently describes how a set of $n$ sampled chromosomes
are related to each other along a (linear) genome of length $L$ \citep{kelleher2016efficient,kelleher2018efficient}.
% These trees are highly correlated,
% a fact which allows them to be stored and processed very efficiently,
% as described in \citet{kelleher2018efficient}.
From the tree sequence can be extracted a sequence of trees,
$\treeseq = (T_1, T_2, \ldots, T_{n(\treeseq)})$
and a sequence of breakpoints $0 = a_0 < a_1 < \cdots < a_{n(T)} = L$,
where $T_k$ describes the genealogical relationships of the samples
over the segment of genome betwen $a_{k-1}$ (inclusive) and $a_k$ (exclusive).
We say that tree $T_k$ \emph{covers} the (half-open) segment $[a_{k-1}, a_k)$,
and call the length of this segment its \emph{span}, denoted $L_k = a_k - a_{k-1}$.
We refer to the branches in each tree using the most recent node,
so for instance the branch describing a relationship where $u$ is the child of $v$
is associated with $u$.
The \emph{length} of this branch is the difference in birth times of $v$ and $u$,
and is denoted $t_T(u)$ (dropping $T$ if it is clear from context).
Note that although the same parent-child relationship may exist across many adjacent trees
(this is called an ``edge''),
rearrangements of genealogical relationships
can cause the precise set of samples that inherit from that edge to differ across
trees due to recombination.

@@

Perhaps the most natural way to summarise weights on the tree
is simply to examine the values at each node.
This would allow us, for instance, to
count the number of samples that inherit from each node in the tree.
Averaged across trees,
this tells us what proportion of the sample's genomes were inherited from each node.
Motivated by this, we define the
\textbf{node statistic} for node $u$
associated with summary function $f()$ and sample weights $\iw$
to be the sum of $f()$ applied to the weight of the subtree inheriting from node $u$
and to the remaining weight,
averaged across the genome:
\begin{align}
    \node(f, \iw)_u
    =
    \frac{1}{L} \sum_{k=1}^{n(\treeseq)} L_k \left( f(\nw_{T_k}(u)) + f(\tiw - \nw_{T_k}(u)) \right).
\end{align}
The weight $\nw_{T_k}(u)$ is the subtree weight of node $u$ in the tree $T_k$,
and $\tiw - \nw_{T_k}(u)$ is the sum of all weights \emph{not} in the subtree of $u$ in $T_k$.
% i.e., the length of the segment of genome that the tree extends for.
% If this is computed over only a ``window'' of the genome
% then $L_k$ is replaced by the length of the portion of that window
% that $T$ extends for.

\paragraph{Polarisation}
The node statistic defined above is appropriate for statistics such as nucleotide diversity, which is a function
of all alleles at given variable site.  For other statistics, one is only concerned with the derived state(s)
at a site, and we may defined a \emph{polarised} node statistic without the second term from our previous definition:

@@

\begin{example}[Ancestry proportions] \label{ex:ancestry_props}
    If $\iw$ are the indicator weights of the set $S$ of $n$ samples,
    and then $\iw_u / n$ is the proportion of the samples in $S$ that inherit from $u$.
    Therefore, if $f(x) = x / n$,
    then $\nodep(f, \iw)_u$ is the proportion of the genomes of $S$
    that are inherited from (ancestor) $u$.
\end{example}

@@

Now we arrive at our primary goal, which is to compute statistics from \emph{genomes}
using this framework.
To do this, we assume that the genetic variation data is embedded in a tree sequence,
but the trees are used only for efficiency --
the results will not depend on the trees in any way
(with the only exception being if the ancestral allele is assumed to be known, as discussed below).

@@

Having defined a general framework, we may now give concrete examples of functions related to common summary statistics.

\begin{example}[Nucleotide diversity] \label{ex:site_diversity}
    The nucleotide diversity of a sample is the average number of differences between all haplotypes, which is
    equivalent to the sum of heterozygosity at all positions.
    To calculate this statistic,
    % let $\iw_u = 1$ for $u \in S$ and $\iw_v = 0$ otherwise,
    let $\iw = \bone_S$,
    so that $\nw(u)$ gives the number of nodes in $S$ inheriting from $u$,
    and define
    \begin{align*}
        f(x) = \frac{x (n - x)}{n (n-1)} .
    \end{align*}
    Then $\site(f, \iw)_{[a,b)}$ is mean nucleotide diversity in the region between $a$ and $b$.
\end{example}

\begin{example}[Nucleotide divergence] \label{ex:site_divergence}
    Now suppose we want to compute the mean density of nucleotide differences
    \emph{between} two nonoverlapping groups of samples, $S_1$ and $S_2$,
    with $n_1$ and $n_2$ samples, respectively.
    As before,
    let $\iw_j = \bone_{S_j}$,
    so that $\nw_{j}(u)$ gives the number of nodes in $S_j$ inheriting from $u$.
    Then we define
    \begin{align*}
        f(x_1, x_2) = \frac{x_1 (n_2 - x_2)}{n_1 n_2} .
    \end{align*}
    Then $\site(f, \iw)_{[a,b)}$ is mean nucleotide divergence between the two groups
    in the region between $a$ and $b$.
\end{example}

@@

\begin{example}[Segregating sites] \label{ex:segregating_sites}
    Again, let $\iw = \bone_{S}$ for a group of $n$ sequences,
    and now define
    \begin{align*}
        f(x) = \begin{cases}
            1 - \frac{x}{n} \qquad &\text{if } x > 0 \\
            0 \qquad &\text{otherwise} .
        \end{cases}
    \end{align*}
    In computing $\site(f, \iw)_{[a,b)}$, the function $f()$ is given the number of samples in $S$
    that inherit each allele.
    A site with $k$ distinct alleles that segregate in $S$ with proportions $p_1 + \cdots + p_k = 1$
    contributes $f(p_1) + \cdots + f(p_k) = (1 - p_1) + \cdots + (1 - p_k) = k - 1$ to the statistic.
    Therefore, $\site(f, \iw)_{[a,b)}$, is the minimum number of derived mutations per unit length,
    which is the density of segregating sites if all sites are biallelic.
\end{example}

\begin{example}[Phenotypic correlations] \label{ex:site_correlations}
    Suppose that for each sample $u$ we have a numeric phenotype, denoted $z(u)$,
    and we want to compute the correlation between this phenotype
    and the genotype at each site in the tree sequence.
    For convenience, suppose $z$ is normalized
    % so that $\sum_u z(u) = 0$ and $\sum_u z(u)^2/(n-1) = 1$.
    to have mean zero and variance 1.
    Then, if $g_j$ is a vector of binary genotypes (so $g_j(u) = 1$ if $u$ carries the derived allele),
    then the covariance of $z$ with $g_j$ is just $\sum_u z(u) g_j(u) = \sum_{u \st g_j(u) = 1} z(u)$,
    i.e., the sum of the phenotypes of all samples carrying the derived allele,
    divided by $(n - 1)$, where $n$ is the number of samples.
    Since the phenotypes sum to zero, this is also equal to
    $- \sum_{u \st g_j(u) = 0} z(u)$.
    If $p_j = \sum_u g_j(u)$ is the derived allele frequency at site $j$,
    then the variance of $g_j$ is $n p_j (1-p_j) / (n-1)$,
    and so the squared correlation can be calculated as a sum across the two alleles:
    \begin{align*}
        r_j^2 =
        \frac{\left( \sum_{u \st g_j(u) = 0} z(u)\right)^2}{2p(1-p)n(n-1)}
        + \frac{\left( \sum_{u \st g_j(u) = 1} z(u)\right)^2}{2p(1-p)n(n-1)}  .
    \end{align*}
    We can compute this as a site statistic by defining $\iw_{1}(u) = z(u)$, and $\iw_{2}(u) = 1/n$,
    and $f(x_1, x_2) = x_1^2 / (2 x_2 (1 - x_2) n (n-1))$:
    then $\site(\iw, f)_j = r_j^2$ is the squared correlation between $z$ and the allele at site $j$.
    Appendix \ref{apx:regression} explains how to extend this to regression with covariates.
\end{example}

%% Check that math:
% n <- 10; z <- rnorm(10); z <- (z - mean(z))/sd(z); g <- rbinom(n, size=1, prob=1/2); p <- mean(g)
% c(cov(z,g), sum(z*g)/(n-1))
% c(cor(g,z), sum(g*z)/sqrt(n*(n-1)*p*(1-p)))
\krt{There is a possible point of confusion here.  So far, weight has been used to describe a property of nodes, and
    then we have a function that operates on those weights.  In this example, the output of that function is also called
a weight.  We may prefer to use a term like ``value" here?}
\begin{example}[Patterson's $f_4$] \label{ex:site_f4}
    Given four disjoint groups of samples, $S_1$, $S_2$, $S_3$, and $S_4$,
    Patterson's $f_4(S_1, S_2; S_3, S_4)$ statistic for an allele with frequency $p_i$ in group $S_i$
    is $(p_1 - p_2)(p_3 - p_4)$ (and then this is typically averaged over sites).
    To rewrite this as a sum over alleles, note that
    $p_1 - p_2 = p_1 (1 - p_2) - (1 - p_1) p_2$,
    and so the statistic counts with positive weight
    alleles that split $S_1$ and $S_3$ from $S_2$ and $S_4$,
    and negative weight ones that split $S_1$ and $S_4$ from $S_2$ and $S_3$.
    Therefore, if as before we
    let $\iw_j = \bone_{S_j}$
    % let $\iw_{ju} = 1$ if $u \in S_j$
    (so that $\iw_{ju}$ tells us the number of samples in $S_j$ descended from $u$),
    and write $n_i$ for the number of samples in $S_i$, and
    \begin{align*}
        f(x_1, x_2, x_3, x_4)
        =
        \frac{x_1}{n_1}
        \frac{x_3}{n_3}
        \left(1 - \frac{x_2}{n_2}\right)
        \left(1 - \frac{x_4}{n_4}\right)
        -
        \left(1 - \frac{x_1}{n_1}\right)
        \left(1 - \frac{x_4}{n_4}\right)
        \frac{x_2}{n_2}
        \frac{x_3}{n_3} ,
    \end{align*}
    then $\site(\iw, f)_{[a,b)}$ is equal to Patterson's $f_4(S_1, S_2; S_3, S_4)$ statistic
    for the region of the genome between $a$ and $b$.
\end{example}

\krt{There is another approach, introduced by Molly Prz.  I'll see if I can dig up a description.  The approach
described here is preferable, though.}
\paragraph{Polyallelic sites}
The statistics defined above map exactly onto those standard in population genetics
when all sites are biallelic.
There is not a consensus in the field
on how to use sites with more than two alleles, but the most common practice is
to reduce to biallelic data, by either discarding other sites
or by marking all non-ancestral alleles as (the same) ``derived'' allele.
We choose to define statistics in a way that still makes formal sense for polyallelic sites,
so that for instance a site with ancestral allele $A$ and derived alleles $C$ and $T$
would have site statistic $f(\aw(A)) + f(\aw(C)) + f(\aw(T))$.
This is a natural defintion in that it agrees with what is obtained by
looking at pairwise haplotype differences.
For instance, the definition of nucleotide divergence in Example \ref{ex:site_divergence}
gives the probability that a randomly chosen sample from each set differ.
This definition makes sense even at polyallelic sites,
and (as we discuss below) has a natural correspondence to statistics related to branch lengths.

@@

Genetic variation is informative about many processes
precisely because it tells us about the underlying patterns of genealogical relatedness.
In other words, the genomes are useful in so far as they tell us about the trees.
In the case where we actually have the trees, or a good proxy for them,
it is natural to summarize them directly, rather than working indirectly with the genotypes.
If we assume that no two mutations occur at the same genomic position --
i.e., the \emph{infinite sites model} --
then there is a natural correspondence between summaries of genotypes and summaries of tree shape.
If mutations occur at a constant rate in time and along the genome,
then the \emph{expected} number of mutations that occur somewhere along a branch of a tree
over some segment of genome
is equal to the mutation rate multiplied by the length of the segment and by the length of the branch.
In other words, the ``area'' of a branch in a tree of the tree sequence,
defined as its span (right minus left endpoint) multiplied by its length (parent time minus child time)
and scaled by the mutation rate,
is equal to the expected number of mutations that will land on it.
If the mutation rate is constant,
then this gives us an isometry --
measuring tree distances in branch lengths versus in numbers of mutations
are equal in expectation, up to a constant.

This makes it natural to define
a statistic of tree shape by summing these expected contributions across its branches.
We define the \textbf{branch statistic} of a single tree $T$ with span $L$
obtained from summary function $f()$
and sample weights $\iw$ with total weight $\tiw = \sum_u \iw(u)$
to be the sum of the area of each branch in each tree
multiplied by the summary function applied to its subtree weight:
\begin{align}
    \branch(f, \iw)_T
    &=
    \frac{1}{L} \sum_{u \in T} t_{u} \left( f(\nw_{T}(u)) + f(\tiw - \nw_{T}(u)) \right)  ,
\end{align}
where $t_{u}$ is the length of the branch ancestral to node $u$ in tree $T$
and $\nw_{T}(u)$ is the total weight of the subtree of $T$ inheriting from $u$ (as defined above).
The term $\tiw - \nw_{T}(u)$ gives the total weight \emph{not} in the subtree of $T$ inheriting from $u$,
because the sum of all weights in the tree is always $\tiw$.
The value $f(\nw(u))$ is the summary value that would be added to a site statistic
if a single mutation occurred on the branch ancestral to $u$,
and $f(\tiw - \nw_{T}(u))$ is the value that would be added due to its complementary allele,
so $\branch(f, \iw)_T$ gives the expected contribution of this branch to $\site(f, \iw)$,
per unit of sequence length that the tree $T$ covers.

@@

@@

\section{Linear regression}

>>>>

Many population genetic summary statistics can be defined using the allele frequency spectrum
of a collection of homologous sequences.
Concretely, write $p_i(a)$ for the frequency of allele $a$ at a site $i$,
and $h(p)$ for the allele frequency spectrum
-- i.e., the number of alleles whose frequency is $p$ in our sample --
and suppose that for some function $f()$, we want to compute $\sum_p h(p) f(p)$.
(For instance, mean genetic diversity is computed with $f(p) = p (1-p)$.)
Since this is a simple sum over sites,
we can compute this by summing over the $L$ loci, as
\begin{align*}
    \sum_p h(p) f(p) = \sum_{i=1}^L \sum_a f(p_i(a)).
\end{align*}
(Note that since the ancestral allele is included, this function is invariant to relabeling of alleles.)
If we know where each mutation has occurred on the genealogy,
we can easily find the frequency of the corresponding allele by counting how many samples
inherit from that mutation in the tree (and accounting for those that subsequently mutate).

This paper presents one way of generalizing of this idea.
Roughly speaking, to compute a statistic we will
propagate ``weights'' additively up each tree,
summarize these weights on each branch with a ``summary function'',
and aggregate these summaries to produce statistics
(e.g., averaged in windows).
For instance, we could calculate how many of a certain set $A$ of samples
inherit from a particular edge in a tree
by (a) assigning each of these samples weight 1 (and other nodes weight zero), then
(b) finding the total weight of all nodes in the subtree inheriting from that edge.
It turns out to be useful to allow more general weights (including negative ones).
Figure \ref{fig:divergence_diagram} shows a simple example of these procedures
of propagating weights up a tree and summarising them,
used to calculate mean sequence divergence over a short sequence spanned by a single tree.

@@

\begin{definition}[Sample and subtree weights] \label{defn:weights}
    The \emph{samples} of a tree sequence are the focal nodes,
    and appear in the trees at every point along the genome.
    A list of \emph{sample weights} $\iw$ assigns a numeric value $\iw(v)$
    to every sample node.
    Given these and a tree $T$,
    the \emph{subtree weight} $\nw_T(u)$ for a node $u$ is the sum of all sample weights
    of every sample node that is descended $u$ in the tree (including $u$, if it is a sample):
    \begin{align*}
        \nw_T(u) = \sum_{v \,:\, v \le_T u} \iw(v) ,
    \end{align*}
    where $v \le_T u$ if $u$ is on the path from $v$ to root in the tree $T$.
    The \emph{total weight} is the sum of the weights over all samples:
    $\tiw = \sum_v \iw(v)$, which is the subtree weight of the root of any tree
    that contains all samples.
\end{definition}

We allow vector-valued weights,
i.e., $\iw(v)$ may be a vector $(\iw_1(v), \ldots, \iw_m(v))$,
so that when summarising we have access to more than one aspect of each subtree.

@@

The requirement that $f(0) = f(\tiw) = 0$ is for consistency,
so that statistics do not depend on portions of the tree either not ancestral to any of the samples or ancestral to all
of them.
% Below, in section XXX, we describe how to efficiently
% maintain a list of the weights of the subtrees inherit from each node
% as we iterate through the tree sequence.

\begin{figure}
    \centering
    \includegraphics{figures/divergence_diagram}
    \caption{
        An example of computing sequence divergence between red nodes and blue nodes
        from seven mutations at distinct sites in a sequence of length 10 using the tree.
        The summary function, $f(a,b) = (a/3)(1-b/2)$ gives the probability that
        an allele found in $a$ of the red nodes and $b$ of the blue nodes is present in
        a randomly chosen red node but not in a randomly chosen blue node.
        \textbf{(a)} Weights are assigned to each subtree by adding together the weights of any children,
        plus the weight of the node itself, if it is a sample.
        \textbf{(b)} In this example, subtree weights $(a,b)$ record how many red nodes ($a$)
        and how many blue nodes ($b$) lie in that subtree.
        \textbf{(c)} The summary function shown is applied to both the weight of the subtree
        and the remaining weight of each mutation,
        so for instance the mutation above the node with weights $(2,1)$
        contributes $f(2,1) + f(1,1) = 1/3 + 1/6 = 1/2$.
        The final value for divergence is the sum of these values divided by 10,
        the total number of sites.
        \label{fig:divergence_diagram}
    }
\end{figure}

@@

Before defining the different classes of statistic,
it will help to establish a little more notation.
A tree sequence efficiently describes how a set of $n$ sampled chromosomes
are related to each other along a (linear) genome of length $L$ \citep{kelleher2016efficient,kelleher2018efficient}.
% These trees are highly correlated,
% a fact which allows them to be stored and processed very efficiently,
% as described in \citet{kelleher2018efficient}.
From the tree sequence can be extracted a sequence of trees,
$\treeseq = (T_1, T_2, \ldots, T_{n(\treeseq)})$
and a sequence of breakpoints $0 = a_0 < a_1 < \cdots < a_{n(T)} = L$,
where $T_k$ describes the genealogical relationships of the samples
over the segment of genome betwen $a_{k-1}$ (inclusive) and $a_k$ (exclusive).
We say that tree $T_k$ \emph{covers} the (half-open) segment $[a_{k-1}, a_k)$,
and call the length of this segment its \emph{span}, denoted $L_k = a_k - a_{k-1}$.
We refer to the branches in each tree using the most recent node,
so for instance a branch between an ancestor $v$ and descendant $u$
is associated with the node $u$.
The \emph{length} of this branch is the difference in birth times of $v$ and $u$,
and is denoted $t_T(u)$ (dropping $T$ if it is clear from context).
Note that although the same parent-child relationship may exist across many adjacent trees
(this is called an ``edge''),
rearrangements of genealogical relationships due to recombination
can cause the precise set of samples that inherit from that edge to differ across trees.

@@

Perhaps the most natural way to summarise weights on the tree
is simply to examine the values at each node.
This would allow us, for instance, to
count the number of samples that inherit from each node in the tree.
Averaged across trees,
this tells us, for each node, what proportion of the sample's genomes were inherited from that node.
Motivated by this, we define the
\textbf{node statistic} for node $u$
associated with summary function $f()$ and sample weights $\iw$
to be the sum of $f()$ applied to the weight of the subtree inheriting from node $u$
and to the remaining weight,
averaged across the genome:
\begin{align}
    \node(f, \iw)_u
    =
    \frac{1}{L} \sum_{k=1}^{n(\treeseq)} L_k \left( f(\nw_{T_k}(u)) + f(\tiw - \nw_{T_k}(u)) \right).
\end{align}
The weight $\nw_{T_k}(u)$ is the subtree weight of node $u$ in the tree $T_k$,
and $\tiw - \nw_{T_k}(u)$ is the sum of all weights \emph{not} in the subtree of $u$ in $T_k$.
% i.e., the length of the segment of genome that the tree extends for.
% If this is computed over only a ``window'' of the genome
% then $L_k$ is replaced by the length of the portion of that window
% that $T$ extends for.

\paragraph{Polarisation}
The node statistic defined above is appropriate for statistics such as nucleotide diversity, which is a function
of all alleles at given variable site.  For other statistics, one is only concerned with the derived state(s)
at a site, and we may define a \emph{polarised} node statistic without the second term from our previous definition:

@@

\begin{example}[Ancestry proportions] \label{ex:ancestry_props}
    If $\iw$ are the indicator weights of the set $S$ of $n$ samples,
    then $\nw(u) / n$ is the proportion of the samples in $S$ that inherit from $u$.
    Therefore, if $f(x) = x / n$,
    then $\nodep(f, \iw)_u$ is the proportion of the genomes of $S$
    that are inherited from (ancestor) $u$.
\end{example}

@@

Now we describe how to compute statistics from \emph{genomes} using this framework.
To do this, we assume that the genetic variation data is embedded in a tree sequence,
but the trees are used only for efficiency --
the results will not depend on the trees in any way
(with the only exception being if the ancestral allele is assumed to be known, as discussed below).

@@

Having defined a general framework, we may now give concrete examples
of how to compute common summary statistics

\begin{example}[Nucleotide diversity] \label{ex:site_diversity}
    The nucleotide diversity of a group $S$ of $n$ samples
    is the average density of differences between pairs of samples,
    or equivalently, the average heterozygosity across positions.
    To calculate this statistic,
    % let $\iw_u = 1$ for $u \in S$ and $\iw_v = 0$ otherwise,
    let $\iw = \bone_S$,
    so that $\nw(u)$ gives the number of nodes in $S$ inheriting from $u$,
    and define
    \begin{align*}
        f(x) = \frac{x (n - x)}{n (n-1)} .
    \end{align*}
    Then $\site(f, \iw)_{[a,b)}$ is mean nucleotide diversity of $S$ in the region between $a$ and $b$.
\end{example}

\begin{example}[Nucleotide divergence] \label{ex:site_divergence}
    Now suppose we want to compute the mean pairwise density of nucleotide differences
    \emph{between} two nonoverlapping groups of samples, $S_1$ and $S_2$,
    with $n_1$ and $n_2$ samples, respectively.
    As before,
    let $\iw_j = \bone_{S_j}$,
    so that $\nw_{j}(u)$ gives the number of nodes in $S_j$ inheriting from $u$,
    and define
    \begin{align*}
        f(x_1, x_2) = \frac{x_1 (n_2 - x_2)}{n_1 n_2} .
    \end{align*}
    Then $\site(f, \iw)_{[a,b)}$ is mean nucleotide divergence between the two groups
    in the region between $a$ and $b$.
\end{example}

@@

\begin{example}[Segregating sites] \label{ex:segregating_sites}
    Again, let $\iw = \bone_{S}$ for a group of $n$ samples,
    and now define
    \begin{align*}
        f(x) = \begin{cases}
            1 - \frac{x}{n} \qquad &\text{if } x > 0 \\
            0 \qquad &\text{otherwise} .
        \end{cases}
    \end{align*}
    In computing $\site(f, \iw)_{[a,b)}$, the function $f()$ is given the number of samples in $S$
    that inherit each allele.
    A site with $k$ distinct alleles that segregate in $S$ with proportions $p_1 + \cdots + p_k = 1$
    contributes $f(p_1) + \cdots + f(p_k) = (1 - p_1) + \cdots + (1 - p_k) = k - 1$ to the statistic.
    Therefore, $\site(f, \iw)_{[a,b)}$, is the minimum number of derived mutations per unit length,
    which is the density of segregating sites if all sites are biallelic.
\end{example}

\begin{example}[Phenotypic correlations] \label{ex:site_correlations}
    Suppose that for each sample $u$ we have a numeric phenotype, denoted $z(u)$,
    and we want to compute the correlation between this phenotype
    and the genotype at each site in the tree sequence.
    For convenience, suppose $z$ is normalized
    % so that $\sum_u z(u) = 0$ and $\sum_u z(u)^2/(n-1) = 1$.
    to have mean zero and variance 1.
    Then, if $g_j$ is a vector of binary genotypes (so $g_j(u) = 1$ if $u$ carries the derived allele),
    then the covariance of $z$ with $g_j$ is just $\sum_u z(u) g_j(u) = \sum_{u \st g_j(u) = 1} z(u)$,
    i.e., the sum of the phenotypes of all samples carrying the derived allele,
    divided by $(n - 1)$, where $n$ is the number of samples.
    Since the phenotypes sum to zero, this is also equal to
    $- \sum_{u \st g_j(u) = 0} z(u)$.
    If $p_j = \sum_u g_j(u)$ is the derived allele frequency at site $j$,
    then the variance of $g_j$ is $n p_j (1-p_j) / (n-1)$,
    and so the squared correlation can be calculated as a sum across the two alleles:
    \begin{align*}
        r_j^2 =
        \frac{\left( \sum_{u \st g_j(u) = 0} z(u)\right)^2}{2p(1-p)n(n-1)}
        + \frac{\left( \sum_{u \st g_j(u) = 1} z(u)\right)^2}{2p(1-p)n(n-1)}  .
    \end{align*}
    We can compute this as a site statistic by defining $\iw_{1}(u) = z(u)$, and $\iw_{2}(u) = 1/n$,
    and $f(x_1, x_2) = x_1^2 / (2 x_2 (1 - x_2) n (n-1))$:
    then $\site(\iw, f)_j = r_j^2$ is the squared correlation between $z$ and the allele at site $j$.
\end{example}

Appendix \ref{apx:regression} explains how to extend the previous example
to obtain the squared coefficient of genotype in the regression of phenotype
against genotype with additional covariates.

%% Check that math:
% n <- 10; z <- rnorm(10); z <- (z - mean(z))/sd(z); g <- rbinom(n, size=1, prob=1/2); p <- mean(g)
% c(cov(z,g), sum(z*g)/(n-1))
% c(cor(g,z), sum(g*z)/sqrt(n*(n-1)*p*(1-p)))
\begin{example}[Patterson's $f_4$] \label{ex:site_f4}
    Given four disjoint groups of samples, $S_1$, $S_2$, $S_3$, and $S_4$,
    Patterson's $f_4(S_1, S_2; S_3, S_4)$ statistic for an allele with frequency $p_i$ in group $S_i$
    is $(p_1 - p_2)(p_3 - p_4)$ (and then this is typically averaged over sites).
    To rewrite this as a sum over alleles, note that
    $p_1 - p_2 = p_1 (1 - p_2) - (1 - p_1) p_2$,
    and so the statistic counts with positive value
    alleles that split $S_1$ and $S_3$ from $S_2$ and $S_4$,
    and negative value ones that split $S_1$ and $S_4$ from $S_2$ and $S_3$.
    Therefore, if as before we
    let $\iw_j = \bone_{S_j}$
    % let $\iw_{ju} = 1$ if $u \in S_j$
    (so that $\iw_{ju}$ tells us the number of samples in $S_j$ descended from $u$),
    and write $n_i$ for the number of samples in $S_i$, and
    \begin{align*}
        f(x_1, x_2, x_3, x_4)
        =
        \frac{x_1}{n_1}
        \left(1 - \frac{x_2}{n_2}\right)
        \frac{x_3}{n_3}
        \left(1 - \frac{x_4}{n_4}\right)
        -
        \left(1 - \frac{x_1}{n_1}\right)
        \frac{x_2}{n_2}
        \frac{x_3}{n_3}
        \left(1 - \frac{x_4}{n_4}\right),
    \end{align*}
    then $\site(\iw, f)_{[a,b)}$ is equal to Patterson's $f_4(S_1, S_2; S_3, S_4)$ statistic
    for the region of the genome between $a$ and $b$.
\end{example}

\krt{There is another approach, introduced by Molly Prz.  I'll see if I can dig up a description.  The approach
described here is preferable, though.}

\paragraph{Polyallelic sites}
The statistics defined above map exactly onto those standard in population genetics
when all sites are biallelic.
There is not a consensus in the field
on how to use sites with more than two alleles, but the most common practice is
to reduce to biallelic data, by either discarding other sites
or by marking all non-ancestral alleles as (the same) ``derived'' allele.
We choose to define statistics in a way that still makes formal sense for polyallelic sites,
so that for instance a site with ancestral allele $A$ and derived alleles $C$ and $T$
would have site statistic $f(\aw(A)) + f(\aw(C)) + f(\aw(T))$.
This is not the only choice, but is natural because it agrees with what is obtained by
looking at pairwise haplotype differences.
For instance, the definition of nucleotide divergence in Example \ref{ex:site_divergence}
gives the probability that a randomly chosen sample from each set differ.
This definition makes sense even at polyallelic sites,
and (as we discuss below) has a natural correspondence to statistics related to branch lengths.

@@

Genetic variation is informative about many processes
precisely because it tells us about the underlying patterns of genealogical relatedness.
In other words, often the genomes are most useful in so far as they tell us about the trees.
In the case where we actually have the trees, or a good proxy for them,
it is natural to summarize them directly, rather than working indirectly with the genotypes.
If we assume that no two mutations occur at the same genomic position --
i.e., the \emph{infinite sites model} --
then there is a natural correspondence between summaries of genotypes and summaries of tree shape.
If mutations occur at a constant rate in time and along the genome,
then the \emph{expected} number of mutations that occur somewhere along a branch of a tree
over some segment of genome
is equal to the mutation rate multiplied by the length of the segment and by the length of the branch.
In other words, the ``area'' of a branch in a tree of the tree sequence,
defined as its span (right minus left endpoint) multiplied by its length (parent time minus child time)
and scaled by the mutation rate,
is equal to the expected number of mutations that will land on it.
If the mutation rate is constant,
then this gives us an isometry --
measuring tree distances in branch lengths versus in numbers of mutations
are equal in expectation, up to a constant.

This makes it natural to define
a statistic of tree shape by summing these expected contributions across its branches.
We define the \textbf{branch statistic} of a single tree $T$
obtained from summary function $f()$
and sample weights $\iw$ with total weight $\tiw = \sum_u \iw(u)$
to be the sum of the length of each branch in each tree
multiplied by the summary function applied to its subtree weight and its complement:
\begin{align}
    \branch(f, \iw)_T
    &=
    \sum_{u \in T} t_{u} \left( f(\nw_{T}(u)) + f(\tiw - \nw_{T}(u)) \right)  ,
\end{align}
where $t_{u}$ is the length of the branch ancestral to node $u$ in tree $T$
and $\nw_{T}(u)$ is the total weight of the subtree of $T$ inheriting from $u$ (as defined above).
The term $\tiw - \nw_{T}(u)$ gives the total weight \emph{not} in the subtree of $T$ inheriting from $u$,
because the sum of all weights in the tree is always $\tiw$.
The value $f(\nw(u))$ is the summary value that would be added to a site statistic
if a single mutation occurred on the branch ancestral to $u$,
and $f(\tiw - \nw_{T}(u))$ is the value that would be added due to its complementary allele,
so $\branch(f, \iw)_T$ gives the expected contribution of the tree $T$ to $\site(f, \iw)$,
per unit of sequence length that the tree covers.

@@

\plr{not sure if we need this section}

@@

\section{Linear regression}
\label{apx:regression}
